Local Type Inference for Polarised System F with Existentials

### 2023-10-17 oopsla24/tex/01-introduction.tex
some more intro text

>>>>

Over the last half-century, there has been considerable work on developing type inference algorithms for programming languages,  mostly aimed at solving the problem of \emph{type polymorphism}.

That is, in pure polymorphic lambda calculus (system F), the polymorphic type $\forall a.\,A$ has a big lambda $\Lambda a.\,e$ as an introduction form, and an explicit type application $e\,[A]$ as an elimination form. This is an extremely simple and compact type system, whose rules fit on a single page, but whose semantics are sophisticated enough to model things like parametricity and representation independence. However, System F by itself is unwieldy as a programming language. The fact that the universal type $\forall a.\,A$ has explicit eliminations means that programs written using polymorphic types will need to be stuffed to the gills with type annotations explaining how and when to instantiate the quantifiers.

Therefore, most work on type inference has been aimed at handling type instantiations implicitly -- we want to be able to use a polymorphic function like $\mathsf{len} : \forall a.\, \mathsf{List}\,a \to \mathsf{int}$ at any concrete list type without explicitly instantiating the quantifier in $\mathsf{len}$'s type. That is, we want to write $\mathsf{len}\,[1,2,3]$ instead of writing $\mathsf{len}\,[\mathsf{int}]\,[1,2,3]$.

The most famous of the algorithms for solving these constraints is the Damas-Hindley-Milner algorithm. The idea is that type instantiation induces a subtype ordering: the type $\forall a.\,A$ is a subtype of all of its instantiations. So we wish to be free to use the same function $\mathsf{len}$ at many different types such as
$\mathsf{List}\,\mathsf{int} \to \mathsf{int}$,
$\mathsf{List}\,\mathsf{bool} \to \mathsf{int}$,
$\mathsf{List}\,(\mathsf{int \times bool}) \to \mathsf{int}$, and so on. However, the subtype relation is nondeterministic: it tells us that whenever we see a polymorphic type $\forall a.\, A$, we know it is a subtype of \emph{any} of its instantiations. To turn this into an algorithm, we have to actually make some choices, and DHM works by using \emph{unification}. Whenever we would have had to introduce a particular concrete type in the specification, the DHM algorithm introduces a unification variable, and incrementally instantiates this variable as more and more type constraints are observed.

However, the universal quantifier is not the only quantifier! Dual to the universal quantifer $\forall$ is the existential quantifier $\exists$. Even though existential types have an equal logical status to universal types, they have been studied much less frequently in the literature. The most widely-used algorithm for handling existentials is the algorithm of Odersky and Laufer. This algorithm treats existentials in a second-class fashion: they are not explicit connectives of the logic, but rather are tied to datatype declarations. As a result, packing and unpacking existential types is tied to constructor introductions and constructor eliminations in pattern matches.
